---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

1. Load the data
The code below creates a "coursera_reproducible_research" directory within user's current working directory, downloads the datafile directly from the source, unzips the file and reads it in using the data.table package. The first few entries of the dataset are presented in the output below.

```{r}
library("data.table")

if (!file.exists("coursera_reproducible_research")){dir.create("coursera_reproducible_research")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./coursera_reproducible_research/repdata%2Fdata%2Factivity.zip", method = "curl")
unzip(zipfile="./coursera_reproducible_research/repdata%2Fdata%2Factivity.zip",
    exdir="coursera_reproducible_research")
activity_data<-read.csv("./coursera_reproducible_research/activity.csv", header=TRUE)
head(activity_data)
```

2. Process/transform the data (if necessary) into a format suitable for your analysis.

Other than having NAs, in the steps column, the data appears to be clean. Date varaible is stored as a factor so let's change it to the date format.

```{r}
activity_data$date<-as.Date(as.character(activity_data$date),"%Y-%m-%d")

head(activity_data, n=3)
```
## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r}
total_per_day<-with(activity_data, tapply(steps, date, sum, na.rm=TRUE))
head(total_per_day)
```

2. Make a histogram of the total number of steps taken each day

```{r}
library(graphics)
for_hist <- hist(total_per_day, plot = FALSE)
plot(for_hist, labels = TRUE, col = "Midnight Blue", main="Total number of Steps per day", xlab="Number of steps taken")
```

3. Calculate and report the mean and median of the total number of steps taken per day.

```{r}
mean_per_day<-mean(total_per_day, na.rm=TRUE)
#print(paste("Mean number of steps per day is", round(mean_per_day,2)))
#print(paste("Median number of steps per day is", round(median(total_per_day),2)))
median_per_day<-median(total_per_day, na.rm =TRUE)
```
The mean and median values are
```{r showtable, results="asis"}
#install.packages("xtable")
library(xtable)
results_table<- cbind(mean_per_day,median_per_day)
colnames(results_table)<-c("Mean", "Median")
print(results_table, type="html")
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
daily_activity_average<-by(activity_data$steps,activity_data$interval,mean, na.rm=TRUE)

plot(daily_activity_average, type="l", main="Activity Pattern on an Average Day", xlab="5-minute Intervals starting at 0:00 and ending at 23:55", ylab="Number of Steps on a Typical Day")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_steps<-max(daily_activity_average)
max_step_index<-which(daily_activity_average==max_steps)
max_interval<-names(daily_activity_average[max_step_index])
print(paste("The maximum number of steps is ", round(max_steps,0), " and occurs in the interval ", max_interval ))
```


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missing_values<-sum(is.na(activity_data))
print(paste("There are ", missing_values, " missing values"))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}
#subset the original data set so that only rows with misisng values are in the new dataset
activity_data_missing<-activity_data[is.na(activity_data),]
summary(activity_data_missing)
# the only misisng data is in the steps column

# General Strategy: extract the interval for each missing values and impute median for that interval

# compute median for each interval
daily_activity_median<-with(activity_data, tapply(steps,interval,median, na.rm=TRUE))

# impute values

for (i in 1: dim(activity_data_missing)[1]){
  # read off the interval for this row
  current_interval<- activity_data_missing$interval[i]
  # get median value for that interval
  median_steps<-daily_activity_median[names(daily_activity_median)%in%     as.character(current_interval) ]
  activity_data_missing$steps[i]<-median_steps
}
```


3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# remove NAs from the original dataset and impute previously-computed values back in

complete_cases_data<-activity_data[complete.cases(activity_data), ]
head(newdata)

final_data_set<-rbind(complete_cases_data, activity_data_missing)
head(final_data_set)
# check that dimesions match
dim(activity_data)[1]
dim(complete_cases_data)[1] + dim(activity_data_missing)[1]
dim(final_data_set)[1]

# Everything Matches. Good
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
#compute totals per day
total_Steps<-with(final_data_set, tapply(steps, date,sum))
library(graphics)
for_hist <- hist(total_Steps, plot = FALSE)
plot(for_hist, labels = TRUE, col = "Midnight Blue", main="Total number of Steps per day: Modified Data", xlab="Number of steps taken")
```
The mean and median values are
```{r showtable, results="asis"}
#install.packages("xtable")
library(xtable)
daily_mean<-mean(total_Steps)
daily_median<-median(total_Steps)
mean_and_medain<- cbind(daily_mean,daily_median)
colnames(mean_and_medain)<-c("Mean", "Median")
print(mean_and_medain, type="html")
```

There is a difference in the mean values but the median value has been preserved because I imputed with the median. 
```{r showtable, results="asis"}
#install.packages("xtable")
library(xtable)
summary_table<- rbind(results_table,mean_and_medain)
rownames(summary_table)<-c("Original Data", "With Imputed Values")
print(summary_table, type="html")
```
## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
final_data_set$date<-as.POSIXct(final_data_set$date, format = "%Y-%m-%d")
final_data_set$dow<- weekdays(final_data_set$date)
weekday<-c("Monday","Tuesday","Wednesday","Thursday","Friday")
final_data_set$business_weekend<-NA
final_data_set[final_data_set$dow%in%weekday,"business_weekend"]<-"weekday"
final_data_set[is.na(final_data_set$business_weekend),"business_weekend"]<-"weekend"
final_data_set$business_weekend<-as.factor(final_data_set$business_weekend)
```
2. Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
weekend_data<-final_data_set[final_data_set$business_weekend=="weekend",]
weekday_data<-final_data_set[final_data_set$business_weekend=="weekday",]

weekend_averages<-with(weekend_data, tapply(steps, interval, mean))

weekday_averages<-with(weekday_data, tapply(steps, interval, mean))


data<-rbind(weekend_averages,weekday_averages)
data2<-t(data)
data2<-as.data.frame(unlist(data2))
data2$interval<-row.names(data2)
reshaped_data<-melt(data2, id.vars=c("interval"),measure.vars=c("weekend_averages", "weekday_averages"), variable.name="business_weekend",value.name="steps")
head(reshaped_data)

library(ggplot2)

ggplot(reshaped_data , aes(interval, steps)) + facet_wrap(vars("business_weekend"))+geom_line() + labs(title = "Average Daily Steps by Weekend/Weekday", x = "Interval", y = "Number of Steps") 
```


